\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-28 paper/algorithm.tex
paper: pass through imputation methods section

To reduce the search space, we only consider the minimal imputation, the maximal imputation, and the minimal drop. The minimal imputation (resp. drop) only imputes (resp. drops) the columns required by the relational operator immediately following the imputation. The maximal imputation imputes all of the attributes in the relation which are used by other operators in the query plan or are visible in the query output.

@@

The planner operates as follows.
First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan (\Cref{lst:line:attr-start}-\Cref{lst:line:attr-end}).
This set will be used to determine which attributes are imputed.
Then, it constructs a plan cache. For each table used in the query, any available selections are pushed to the leaves and
a search over imputation operations produces various possible plans, which are added to the cache (\Cref{lst:line:sel}). If no selections are available, a 
simple scan is added to the plan cache (\Cref{lst:line:scan}). Join order is jointly optimized with imputation placements and a set of
plans encompassing all necessary tables is produced (\Cref{lst:line:join}). This set is now extended for any necessary grouping and aggregation
operations (\Cref{lst:line:group}). After a similar step has been taken for projections, the planner now contains the plan frontier (i.e. the
best possible set from which to pick our final plan). The final step in the planner is to find all plans that are
$\alpha$-bound and return the one that has the lowest runtime: the $\alpha$-bound optimal plan.

The join optimization is based on the algorithm used in System R\cite{blasgen1981system},
but rather than select the best plan at every point, we use our plan cache, which keeps track of the tables joined and the dirty-set specific Pareto frontiers.
The output of the join optimizer is the set of plans which satisfy all join predicates and lie along their respective frontier.

@@

We design our system such that any imputation strategy can be plugged in with minimal effort
and without changes to the optimizer, so the strategy can in principle be targeted to a
specific domain. The current version of the system uses a general-purpose imputation
strategy based on chained-equation classification and regression trees (CE-CART). Chained
equation imputation methods \cite{vanbuuren2011mice} (sometimes called \textit{iterative
regression} \cite{gelman2006data}) impute multiple missing attributes by
iteratively estimating predictive models of one missing attribute conditional on all
complete attributes and the other missing attributes. The individual predictive models can
be customized by the analyst to the problem at hand. Decision tree algorithms, like CART,
are found to be effective \cite{akande2015empirical} in empirical studies in general-purpose
routines and are widely used in epidemiological domains \cite{burgette2010multiple}. Though
we provide quantitative results for the CE-CART implementation, ImputeDB is designed to be
agnostic to the choice of imputation algorithm, and additional algorithms --- with
associated time and loss functions --- could be inserted without any change in design.

@@

Our algorithm proceeds by iteratively fitting regression trees to a subset of the data and
the target imputation columns. In each iteration, the missing values of a column are
replaced with newly imputed values. With each epoch, the quality of the imputation improves
as values progressively reflect more accurate relationships amongst attributes. The
algorithm terminates when convergence is achieved (i.e. the imputed values do not change across
epochs) or a fixed number of epochs is reached.

@@

The addition of imputation operators increases the number of plans exponentially, as an imputation may be placed before any of the relational operators.
We restrict imputations to two classes: those that impute only attributes used in the operator and those that impute all the attributes that are needed downstream in the query.
By doing so we limit the number of imputations at any point to four: \textit{Drop} or \textit{Impute} over the attributes used in the operator or over all the downstream attributes.
This modification increases the complexity of our planning algorithm to $O(16^J)$.\todo{I think this is right, but should check}

@@

Finally, we note that for the queries we have examined, this exponential blowup does not affect the practical performance of our optimizer.
Recall that the planner maintains different plans for different dirty sets, keeping only those that are not dominated by other plans with the same dirty set.
So in many cases we can drop some of these intermediate plans generated at each operator.
The worst case complexity only occurs if the dirty sets tracked are distinct through the entire planning phase.
Planning times are discussed in more detail in \Cref{sec:results}.

>>>>

To reduce the search space, we only consider the minimal imputation, the maximal imputation, and the minimal drop. The minimal imputation (resp.\ drop) only imputes (resp.\ drops) the columns required by the relational operator immediately following the imputation. The maximal imputation imputes all of the attributes in the relation which are used by other operators in the query plan or are visible in the query output.

@@

The planner operates as follows.
First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan (\Cref{lst:line:attr-start}--\Cref{lst:line:attr-end}).
This set will be used to determine which attributes are imputed.
Then, it constructs a plan cache. For each table used in the query, any available selections are pushed to the leaves and
a search over imputation operations produces various possible plans, which are added to the cache (\Cref{lst:line:sel}). If no selections are available, a 
simple scan is added to the plan cache (\Cref{lst:line:scan}). Join order is jointly optimized with imputation placements and a set of
plans encompassing all necessary tables is produced (\Cref{lst:line:join}). This set is now extended for any necessary grouping and aggregation
operations (\Cref{lst:line:group}). After a similar step has been taken for projections, the planner now contains the plan frontier (i.e.\ the
best possible set from which to pick our final plan). The final step in the planner is to find all plans that are
$\alpha$-bound and return the one that has the lowest runtime: the $\alpha$-bound optimal plan.

The join optimization is based on the algorithm used in System R~\cite{blasgen1981system},
but rather than select the best plan at every point, we use our plan cache, which keeps track of the tables joined and the dirty-set specific Pareto frontiers.
The output of the join optimizer is the set of plans which satisfy all join predicates and lie along their respective frontier.

@@

\ProjectName{} uses a general-purpose imputation strategy based on chained-equation classification and regression trees (CE-CART).
Chained equation imputation methods~\cite{vanbuuren2011mice} (sometimes called \textit{iterative regression}~\cite{gelman2006data}) impute multiple missing attributes by iteratively fitting a predictive model of one missing attribute which is conditional on all complete attributes and the other missing attributes.
Decision tree algorithms like CART are found to be effective~\cite{akande2015empirical} and are widely used in epidemiological domains~\cite{burgette2010multiple}.

\ProjectName{} is designed so that any imputation strategy can be plugged in with minimal effort and without changes to the optimizer, simply by adding time and loss functions (\Cref{sec:cost}) which describe the new algorithm.
In principle, new imputation algorithms can be added which are targeted to a specific domain.

@@

Our algorithm proceeds by iteratively fitting regression trees to a subset of the data and
the target imputation columns. In each iteration, the missing values of a column are
replaced with newly imputed values. With each epoch, the quality of the imputation improves
as values progressively reflect more accurate relationships amongst attributes. The
algorithm terminates when convergence is achieved (i.e.\ the imputed values do not change across
epochs) or a fixed number of epochs is reached.

@@

The addition of imputation operators increases the number of plans exponentially, as an imputation may be placed before any of the relational operators.
We restrict imputations to two classes: those that impute only attributes used in the operator and those that impute all the attributes that are needed downstream in the query.
By doing so we limit the number of imputations at any point to four: \textit{Drop} or \textit{Impute} over the attributes used in the operator or over all the downstream attributes.
This modification increases the complexity of our planning algorithm to $O(2^{4J})$.

@@

Finally, we note that for the queries we have examined, this exponential blowup does not affect the practical performance of our optimizer.
Recall that the planner maintains different plans for different dirty sets, keeping only those that are not dominated by other plans with the same dirty set.
So in many cases we can drop some of these intermediate plans generated at each operator.
The worst case complexity only occurs if the dirty sets tracked are distinct through the entire planning phase.
Planning times are discussed in \Cref{sec:results}.
